---
title: "Do Musicians learn? Evidence from Song Popularity"
author: "Dr Preet Deep Singh"
affiliation: "Invest India"
date: "09/11/2021"
output: 
  bookdown::pdf_document2:
    toc: no
    citation_package: natbib
    number_sections: yes
    keep_tex: yes
abstract: "Using a dataset of 5,86,672 songs from Spotify, we evaluate whether artists learn from the response their songs get or not. We find that popularity of songs improves with iterations. We plot whether artists have their biggest hits in their first year or subsequent years."
documentclass: "article"
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: sentence
bibliography: references.bib
link-citations: true
csl: apa.csl
---

```{r setup, message=FALSE, warning=FALSE, header=FALSE, echo=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)
library(tidyverse)
library(ARTofR)
library(stargazer)
library(dplyr)
library(ggplot2)
countrywisecollated <- read_csv("Final database.csv") 
countrywisecollated$ReleaseDate<-as.Date(as.character(countrywisecollated$Release_date))
tracks <- read_csv("tracks.csv")
tracks2 <- read_csv("spotify_tracks.csv") ##This has lyrics
colnames(tracks2)[5]<-"id_artists" ##TO make it similar to TRacks file.
tracks$release_date<-as.Date(as.character(tracks$release_date)) #this is now as per date
tracks$YearRelease<-format(tracks$release_date,format="%Y") #Get year also
artistyearfile2<-read_csv("artistyearfile2.csv")
df5<-read_csv("HighestPerDatePerArt.csv")

```
\section{Introduction}


In a series of papers, I try to examine the following questions
                                                                           .
# What does the average album look like: how many sad, what balance
# Do albums with similar songs do better
# Do albums have multiple hit songs or only 1/2 songs make it ?
# Do artists experiment after a hit or do more of the same? Are they scare or overconfident?
# Do negative songs do better?
# Do winter songs do better?
# When does a artist shine to fame: first or next or last?
# Is the same song a hit in multiple places? What are unique geographic hits?

In this paper, I look at the most popular hits of artists. I examine whether artists get their biggest hits in the first go and spend the rest of their lives trying to match that success or whether artists enter the market, release music, take feedback and improve in order to produce more popular songs. In case artists are learning from feedback, the popularity of their songs would have an upward trend. 

In case artists do not take feedback, their most popular songs would be spread randomly across their careers. For artists who had their most popular song in their first year, the subsequent years would entail a lot of experiments, most of which would fail and eventually they would retire. This means we should see artists with their best hits coming in the first year, dropping off in a couple of years post that. The year/date of the last song by a particular band can be used to check this hypothesis.

To examine learning we regress popularity of the songs on their sequence. We find that sequence has a positive coefficient that is statistically significant while controlling for other song traits. 

Our data spans a total of `r length(unique(tracks$id))` songs by `r length(unique(tracks$id_artists))` across `r length(unique(tracks$YearRelease))` years. This includes `r nrow(filter(tracks,explicit==1))` explicit songs. 



```{r artistyear2, eval=FALSE, echo=FALSE, message=FALSE, warning=FALSE, header=FALSE}
## This is just because we did it once, this is not being evaluated.
artistyearfile2<-data.frame()
globalcollated2<-select(tracks, 1:8,YearRelease)
for (i in 1:length(unique(globalcollated2$id_artists))){
  df1<-filter(globalcollated2,id_artists==unique(globalcollated2$id_artists)[i])
  df1<- df1[order(-df1$popularity),] #Arrange in Descending order
  artistyearfile2[i,1]<-  unique(globalcollated2$id_artists)[i]
  artistyearfile2[i,2]<-  min(df1$YearRelease,na.rm = TRUE) #First song #Easy, direct
  artistyearfile2[i,3]<- min(df1$release_date,na.rm=TRUE)
  artistyearfile2[i,4]<-df2$YearRelease[1]#Top Song Year
  artistyearfile2[i,5]<- df1$release_date[1] #NA hai toh avoid kar do .
  artistyearfile2[i,6]<- max(df1$YearRelease,na.rm = TRUE) #First song #Easy, direct
  artistyearfile2[i,7]<- max(df1$release_date,na.rm=TRUE)
  #}
  print(i)
}
colnames(artistyearfile2)<-c("Artist","Year1","Date1","GoldenYear","GoldenDate","LastYear","LastDate")
```

\section{Data}

\subsection{Summary}

* Total songs
* Total Artists
* Total Albums
* Average Popularity
* Year wise distribution



```{r datasummary1, echo=FALSE, message=FALSE, warning=FALSE, header=FALSE, results='asis'}
## countrywisecollated <- read_csv("Final database.csv") 
## countrywisecollated$ReleaseDate<-as.Date(as.character(countrywisecollated$Release_date))
## countrywisecollated$YearRelease<-format(countrywisecollated$ReleaseDate, format= "%Y")
## globalcollated<-filter(countrywisecollated,Country=="Global")
## globalcollated2<-select(globalcollated, 1:10,YearRelease)

stargazer(as.data.frame(tracks),title="This table shows the summary of the dataset.",summary = TRUE,header=FALSE)

```


Figure \@ref(datasummary2) shows the date wise distribution of songs in our dataset. 
Figure \@ref(fig:datasummary3) shows the year wise distribution of songs in our dataset.  ##IF this is correct, use the same label command in the previous one.
```{r datasummary2, echo=FALSE, message=FALSE, warning=FALSE, header=FALSE, }
library(ggplot2)
datewise<-tracks%>%group_by(release_date)%>%summarise(release_date=release_date,songs=length(release_date))
plot<-ggplot(aes(x=release_date,y=songs),data=datewise)+geom_line() +ylim(0,1000)
print(plot)
# fig.cap="\\label{fig:datasummary2}This graph shows the date wise distribution of songs."
```


```{r datasummary3, echo=FALSE, message=FALSE, warning=FALSE, header=FALSE,fig.cap="\\label{fig:datasummary3}This graph shows the year wise distribution of songs."}
library(ggplot2)
tracks$YearRelease<-format(tracks$release_date,format="%Y")
datewise<-tracks%>%group_by(YearRelease)%>%summarise(release_date=YearRelease,songs=length(YearRelease))
datewise<-datewise[complete.cases(datewise),]
datewise<-unique(datewise)
plot<-ggplot(aes(x=YearRelease,y=songs),data=datewise)+geom_bar(stat="identity") +theme(axis.text.x = element_text(angle=-90))+labs(title = "This graph shows the number of songs per year in our dataset.")
print(plot)

```




```{r artistyear3, echo=FALSE, message=FALSE, warning=FALSE, header=FALSE, results='asis'}
artistyearfile2<-read_csv("artistyearfile2.csv")
artistyearfile2<-artistyearfile2%>%filter(!is.na(Year1)) #At least get the years right
artistyearfile2$GYear<-format(artistyearfile2$GoldenYear,format="%Y") #See if th
artistyearfile2$GYear<-as.numeric(artistyearfile2$GYear)
artistyearfile2$Year1<-as.numeric(artistyearfile2$Year1)
artistyearfile2$HitYearDiff<-artistyearfile2$GYear-artistyearfile2$Year1
artistyearfile2$LastYear<-as.numeric(artistyearfile2$LastYear)
artistyearfile2$LastHitYearDiff<- artistyearfile2$LastYear-artistyearfile2$GYear
#artistyearfile2<-mutate(artistyearfile2, HitDayDiff= GoldenDate-Date1)
artistyearfile2$LastHitDayDiff<-artistyearfile2$LastDate-artistyearfile2$GoldenDate

stargazer(as.data.frame(artistyearfile2),title="This table shows the summary of the dates of first song, biggest hit and last song per artist.")
```

#...............................................................................
We look at the year of the first song by the artist and the year of the most popular song by the artist. For example, if the artist released the first song in 1997, and their most popular song was released in 2001, then we note 2001-1997 = 4 as the time till the first hit. In case the most popular song was released in the same year, this number would be 0.

This analysis allows us to answer the following questions
# Who got it soonest
# How many got it soonest
# How many in 1 year
# Who got it late
# How many after 5 years



\subsection{Regression}
Sequence of songs 
Only the most popular song on any date, incase there were multiple songs on the same date
Only artists with more than 1 song 
Get details of the music
Regression on Popularity on All the factors 
Find that sequence is a major contributor along with Explicit

```{r reg1seqonpopularity, echo=FALSE, results='asis', warning=FALSE, header=FALSE}


df5<-read_csv("HighestPerDatePerArt.csv") 
df6<-filter(df5,seq==2) ##To ensure there is at least one more song and not just the first one
df6<-select(df6,id_artists,name) # Name will tell us which ones are there in this 
df6<-merge(df6,df5,by="id_artists",all.y=TRUE) #This is to get traits of the song
df6<-filter(df6,!is.na(df6$name.x)) #This file now has all artists that have more than 1 song in the database.
# To study role of sequence on popularity
df6<-merge(df6,tracks,by="id", all.x=TRUE,all.y=FALSE)

reg1<-lm(popularity.x~seq+duration_ms.x+danceability+energy+key+loudness+mode+speechiness+acousticness+
     instrumentalness+instrumentalness +valence+tempo+time_signature+explicit.x+YearRelease.x,data=df6)
reg2<-lm(popularity.x~seq+duration_ms.x+danceability+energy+key+loudness+mode+speechiness+acousticness+
     instrumentalness+instrumentalness +valence+tempo+time_signature+explicit.x,data=df6)
stargazer(reg1,reg2,omit="YearRelease.x",omit.labels="Year" ,title="This table shows the results of regression of popularity on sequence and other control variables.",no.space=TRUE,header = FALSE ) ###yes, seq has a positive coefficient

```


\section{Limitations}
Limited dataset: It is possible that there are some songs by certain artists that are not part of the half a million songs I use. In some cases, the biggest hit of a few artists might have existed in a particular year and is not accounted for in this study. However, our dataset is not biased in terms of vintage or recency. Therefore, I do not expect this to be systematically lopsided. 
This data is from Spotify. The popularity of the app has gone up in each of the previous few years. It is possible that songs that were not hits in the first go, in the first geography subsequently became hits. This would not be accounted for in our study.


\section{TODO}
1. The year differences dont seem to be correct.
